Follow-up Samples

### 2017-05-03 submission/manuscript.tex

Building upon this evidence, several human studies have identified
unique signatures of colonic lesions {[}15--20{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}21--23{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}24--26{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
have also noted that the loss of bacteria generally thought to produce
short chain fatty acids, which can suppress inflammation, is associated
with colonic lesions. This suggests that gut bacteria have a role in
tumorigenesis with potential as useful biomarkers for aiding in the
early detection of disease.

Despite advances in understanding the role between the gut microbiota
and colonic tumorigenesis, we still do not understand how treatments
including resection, chemotherapy, and/or radiation affect the
composition of the gut microbiota. If the community drives tumorigenesis
then one would hypothesize that treatment to remove a lesion would
affect the microbiota and risk of recurrence. To test this hypothesis,
we addressed two related questions: Does treatment affect the colonic
microbiota in a predictable manner? If so, does the treatment alter the
community to more closely resemble that of individuals with normal
colons?

@@

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the
irregular distribution of OTUs across individuals limited the
statistical power of the test. To overcome these problems we developed
Random Forest models to identify collections of OTUs that would allow us
to differentiate between pre and post-treatment samples from each of the
diagnosis groups. Because of the relatively small number of subjects in
each group, we restricted our models to only incorporate 10 OTUs to
limit the likelihood that the models would overfit the data. Despite
this restriction, the models performed well (adenoma AUC range = 0.69 -
0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
had little overlap with each other {[}Figure 2{]}. These results support
the earlier community-wide analysis where we observed that the treatment
had an impact on the overall community structure; however, the effect of
treatment was not consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S2).
The OTUs that were incorporated into the adenoma and advanced adenoma
models largely overlapped and those OTUs that were used to classify the
carcinoma samples were largely distinct from those of the other two
models {[}Figure 3A{]}. Among the OTUs that were shared across the three
models were those populations generally considered beneficial to their
host (e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figures 3B{]}. Although many of these OTUs were
also included in the model differentiating between patients with normal
colons and those with carcinoma, this model also included OTUs
affiliated with populations that have previously been associated with
carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}24--26{]} {[}Figure S3{]} with some individuals
showing are marked decrease in relative abundance {[}Figure S4{]}.
Finally, we applied these three models to the pre and post-treatment
samples for each diagnosis group and quantified the change in the
positive probability of the model. A decrease in the positive
probability would indicate that the microbiota more closely resembled
that of a patient with a normal colon. There was no significant change
in the positive probability for the adenoma or advanced adenoma groups
{[}Figure 4{]}. The positive probability for the pre and post-treatment
samples from patients diagnosed with carcinoma significantly decreased
with treatment, suggesting a shift toward a normal microbiota for most
individuals. Only, 6 of the 26 patients (23.08\%) who were diagnosed
with a carcinoma had a higher positive probability after treatment; one
of those was re-diagnosed with carcinoma on the follow up visit. These
results indicate that, although there were changes in the microbiota
associated with treatment, those experienced by patients with carcinoma
after treatment yielded gut bacterial communities of greater similarity
to that of a normal colon.

@@

This study builds upon previous work that demonstrated a role for the
microbiota in tumorigenesis and that microbiome-derived biomarkers can
detect colonic lesions {[}6--10,24--28{]}. Our study focused on
comparing the microbiota of patients diagnosed with adenoma, advanced
adenoma, and carcinoma before and after treatment. For all three groups
of patients, we observed changes in their microbiota. After treatment,
the microbiota of patients with carcinoma changed significantly more
than the other groups. This change resulted in communities that more
closely resembled those of patients with a normal colon. This may
suggest that treatment for carcinoma is not only successful for removing
the carcinoma but also at reducing the associated bacterial communities.
Understanding the effect of treatment on the microbiota of those
diagnosed with carcinomas may have important implications for reducing
disease recurrence. It is intriguing that it may be possible to use
microbiome-based biomarkers to not only predict the presence of lesions
but to assess the risk of recurrence.

Patients diagnosed with adenoma and advanced adenoma, however, did not
experience a shift towards a community structure that resembled those
with normal colons. This may be due to the fundamental differences
between the features of adenomas and advanced adenomas and carcinoma.
Specifically, carcinomas may create an inflammatory milieu that would
impact the structure of the community and removal of that stimulus would
alter said structure. In addition, it is possible that the difference
between the microbiota of patients with adenoma and advanced adenoma and
those with normal colons is subtle. This is supported by the reduced
ability of our models to correctly classify patients with adenomas and
advanced adenomas relative to those diagnosed with carcinomas {[}Figure
S2{]}. Given the irregular distribution of microbiota across patients in
the different diagnosis groups, it is possible that we lacked the
statistical power to adequately characterize the change in the
communities following treatment.

There was a subset of patients (6 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 23.08\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 6 individuals (3 with Stage II and 3 with Stage III) was
similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased or decreased (Wilcoxon Test; P-value = 0.56).
Finally, it is possible that these patients may not have responded to
treatment as well as the other 20 patients diagnosed with carcinoma and
so the microbiota may not have been impacted the same way. Again,
further studies looking at the role of the microbiota in recurrence are
needed to understand the dynamics following treatment.

Our final hypothesis was that the specific type of treatment altered the
structure of the microbiome. The treatment to remove adenomas and
advanced adenomas was either polyp removal or surgical resection whereas
it was surgical resection alone or in combination with chemotherapy or
with chemotherapy and radiation for individuals with carcinoma. Because
chemotherapy and radiation target rapidly growing cells, these
treatments would be more likely to cause a turnover of the colonic
epithelium driving a more significant change in the structure of the
microbiota. Although, we were able to test for an effect across these
specific types of treatment, the number of patients in each treatment
group was relatively small.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}24{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between one and two weeks after initial colonoscopy. The bacterial
community has been shown to normalize back to a pre-colonoscopy
community within this time period {[}29{]}. Our study cohort consisted
of 67 individuals with an initial sample as described and a follow up
sample obtained between 188 - 546 days after treatment of lesion
{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
histopathological review of any biopsies taken. Patients were classified
as having advanced adenoma if they had an adenoma greater than 1 cm,
more than three adenomas of any size, or an adenoma with villous
histology. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}30{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between four sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}30{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}31{]}, screening for
chimeras using UCHIME {[}32{]}, classifying with a naive Bayesian
classifier using the Ribosomal Database Project (RDP){[}33{]}, and
clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm
{[}34{]}. The number of sequences for each sample was rarefied to 10523
to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}35{]} algorithm was
used to create the three models used. The adenoma model classified
normal versus adenoma, the advanced adenoma model classified normal
versus advanced adenoma, and the carcinoma model classified normal
versus carcinoma. The total number of individuals in the treatment
models was 67 individuals. There were a total of 22 individuals in the
treatment adenoma model, 19 individuals in the treatment advanced
adenoma model, and 26 individuals in the treatment carcinoma model
{[}Table 1{]}. The total number of individuals in the normal versus
diagnosis models was 423 individuals {[}Table 2{]}. There were a total
of 239 individuals in the adenoma model, 262 individuals in the advanced
adenoma model, and 266 individuals in the carcinoma model {[}Table 2{]}.
Each model was then applied to our 67-person cohort {[}Table 1{]} which
assesed the prediction of pre-treatment adenoma (adenoma n = 22 and
disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
disease free n = 22), pre-treatment advanced adenoma (advanced adenoma n
= 19 and disease free n = 0 ) versus post-treatment advanced adenoma
(advanced adenoma n = 0 and disease free n = 19), pre-treatment
carcinoma (carinoma n = 26 and disease free n = 0) versus post-treatment
carcinoma (carcinoma n = 1 and disease free n = 25).

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}36{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre and post-treatment samples utilized a paired
Wilcoxon ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}37{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtaining the median
MDA from the 100, 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\begin{longtable}[]{@{}lccc@{}}
\toprule
& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
\midrule
\endhead
n & 22 & 19 & 26\tabularnewline
Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
12.9\tabularnewline
Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
BMI (Mean ± SD) & 26.86 ± 3.9 & 25.80 ± 4.7 & 28.63 ± 7.2\tabularnewline
Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
\bottomrule
\end{longtable}

@@

\textbf{Figure S1: Distribution of P-values from Paired Wilcoxson
Analysis of All OTUs Before and After Treatment}

\textbf{Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and
Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
represent the range of values of a 100 different 80/20 splits of the
test set data and the dark green line represents the model using 100\%
of the data set and what was used for subsequent classification. B)
Advanced Adenoma ROC curve: The light yellow shaded areas represent the
range of values of a 100 different 80/20 splits of the test set data and
the dark yellow line represents the model using 100\% of the data set
and what was used for subsequent classification. C) Carcinoma ROC curve:
The light red shaded areas represent the range of values of a 100
different 80/20 splits of the test set data and the dark red line
represents the model using 100\% of the data set and what was used for
subsequent classification.

\textbf{Figure S3: Summary of Important OTUs for the Adenoma, Advanced
Adenoma, and Carcinoma Models.} A) MDA of the most important variables
in the adenoma model. The dark green point represents the mean and the
lighter green points are the value of each of the 100 different runs. B)
Summary of Important Variables in the advanced adenoma model. MDA of the
most important variables in the SRN model. The dark yellow point
represents the mean and the lighter yellow points are the value of each
of the 100 different runs. C) MDA of the most important variables in the
carcinoma model. The dark red point represents the mean and the lighter
red points are the value of each of the 100 different runs.

@@

\hypertarget{ref-yu_metagenomic_2017}{}
27. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
analysis of faecal microbiome as a tool towards targeted non-invasive
biomarkers for colorectal cancer. Gut. 2017;66:70--8.

\hypertarget{ref-warren_co-occurrence_2013}{}
28. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
Microbiome. 2013;1:16.

\hypertarget{ref-obrien_impact_2013}{}
29. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

\hypertarget{ref-kozich_development_2013}{}
30. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology. 2013;79:5112--20.

\hypertarget{ref-pruesse_silva_2007}{}
31. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
SILVA: A comprehensive online resource for quality checked and aligned
ribosomal RNA sequence data compatible with ARB. Nucleic Acids Research.
2007;35:7188--96.

\hypertarget{ref-edgar_uchime_2011}{}
32. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
sensitivity and speed of chimera detection. Bioinformatics (Oxford,
England). 2011;27:2194--200.

\hypertarget{ref-wang_naive_2007}{}
33. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
for rapid assignment of rRNA sequences into the new bacterial taxonomy.
Applied and Environmental Microbiology. 2007;73:5261--7.

\hypertarget{ref-schloss_assessing_2011}{}
34. Schloss PD, Westcott SL. Assessing and improving methods used in
operational taxonomic unit-based approaches for 16S rRNA gene sequence
analysis. Applied and Environmental Microbiology. 2011;77:3219--26.

\hypertarget{ref-breiman_random_2001}{}
35. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
{[}cited 2013 Feb 7{]};45:5--32. Available from:
\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}

\hypertarget{ref-anderson_permanova_2013}{}
36. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
the face of heterogeneous dispersions: What null hypothesis are you
testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
5{]};83:557--74. Available from:
\url{http://doi.wiley.com/10.1890/12-2010.1}

\hypertarget{ref-benjamini_controlling_1995}{}
37. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.

>>>>

Building upon this evidence, several human studies have identified
unique signatures of colonic lesions {[}15--20{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}21--23{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}24--26{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
have also noted that the loss of bacteria generally thought to produce
short chain fatty acids, which can suppress inflammation, is associated
with colonic lesions. This suggests that gut bacteria have a role in
tumorigenesis with potential as useful biomarkers for aiding in the
early detection of disease {[}21--26{]}.

Despite advances in understanding the role between the gut microbiota
and colonic tumorigenesis, we still do not understand how treatments
including resection, chemotherapy, and/or radiation affect the
composition of the gut microbiota. If the microbial community drives
tumorigenesis then one would hypothesize that treatment to remove a
lesion would affect the microbiota and risk of recurrence. To test this
hypothesis, we addressed two related questions: Does treatment affect
the colonic microbiota in a predictable manner? If so, does the
treatment alter the community to more closely resemble that of
individuals with normal colons?

@@

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value NA
0.05). It is likely that high inter-individual variation and the
irregular distribution of OTUs across individuals limited the
statistical power of the test. To overcome these problems we developed
Random Forest models to identify collections of OTUs that would allow us
to differentiate between pre and post-treatment samples from each of the
diagnosis groups. Because of the relatively small number of subjects in
each group, we restricted our models to only incorporate 10 OTUs to
limit the likelihood that the models would overfit the data. Despite
this restriction, the models performed well (adenoma AUC range = 0.69 -
0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
had little overlap with each other {[}Figure 2{]}. These results support
the earlier community-wide analysis where we observed that the treatment
had an impact on the overall community structure; however, the effect of
treatment was not consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed well (adenoma AUC range =0.62 - 0.72, advanced adenoma
AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S1).
The OTUs that were incorporated into the adenoma and advanced adenoma
models largely overlapped and those OTUs that were used to classify the
carcinoma samples were largely distinct from those of the other two
models {[}Figure 3A{]}. Among the OTUs that were shared across the three
models were those populations generally considered beneficial to their
host (e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figures 3B{]}. Although many of these OTUs were
also included in the model differentiating between patients with normal
colons and those with carcinoma, this model also included OTUs
affiliated with populations that have previously been associated with
carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
showing are marked decrease in relative abundance {[}Figure S3{]}.
Finally, we applied these three models to the pre and post-treatment
samples for each diagnosis group and quantified the change in the
positive probability of the model. A decrease in the positive
probability would indicate that the microbiota more closely resembled
that of a patient with a normal colon. There was no significant change
in the positive probability for the adenoma or advanced adenoma groups
{[}Figure 4{]}. The positive probability for the pre and post-treatment
samples from patients diagnosed with carcinoma significantly decreased
with treatment, suggesting a shift toward a normal microbiota for most
individuals. Only, 6 of the 26 patients (23.08\%) who were diagnosed
with a carcinoma had a higher positive probability after treatment; one
of those was re-diagnosed with carcinoma on the follow up visit. These
results indicate that, although there were changes in the microbiota
associated with treatment, those experienced by patients with carcinoma
after treatment yielded gut bacterial communities of greater similarity
to that of a normal colon.

@@

Our study focused on comparing the microbiota of patients diagnosed with
adenoma, advanced adenoma, and carcinoma before and after treatment. For
all three groups of patients, we observed changes in their microbiota.
After treatment, the microbiota of patients with carcinoma changed
significantly more than the other groups. This change resulted in
communities that more closely resembled those of patients with a normal
colon. This may suggest that treatment for carcinoma is not only
successful for removing the carcinoma but also at reducing the
associated bacterial communities. Understanding the effect of treatment
on the microbiota of those diagnosed with carcinomas may have important
implications for reducing disease recurrence. It is intriguing that it
may be possible to use microbiome-based biomarkers to not only predict
the presence of lesions but to assess the risk of recurrence.

Patients diagnosed with adenoma and advanced adenoma, however, did not
experience a shift towards a community structure that resembled those
with normal colons. This may be due to the fundamental differences
between the features of adenomas and advanced adenomas and carcinoma.
Specifically, carcinomas may create an inflammatory milieu that would
impact the structure of the community and removal of that stimulus would
alter said structure. In addition, it is possible that the difference
between the microbiota of patients with adenoma and advanced adenoma and
those with normal colons is subtle. This is supported by the reduced
ability of our models to correctly classify patients with adenomas and
advanced adenomas relative to those diagnosed with carcinomas {[}Figure
S1{]}. Given the irregular distribution of microbiota across patients in
the different diagnosis groups, it is possible that we lacked the
statistical power to adequately characterize the change in the
communities following treatment.

There was a subset of patients (6 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 23.08\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 6 individuals (3 with Stage II and 3 with Stage III) was
similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased (382 \(\pm\) 116 days) or decreased (341 \(\pm\)
98 days) (Wilcoxon Test; P-value = 0.56). Finally, it is possible that
these patients may not have responded to treatment as well as the other
20 patients diagnosed with carcinoma and so the microbiota may not have
been impacted the same way. Again, further studies looking at the role
of the microbiota in recurrence are needed to understand the dynamics
following treatment.

Our final hypothesis was that the specific type of treatment altered the
structure of the microbiome. The treatment to remove adenomas and
advanced adenomas was either polyp removal or surgical resection whereas
it was surgical resection alone or in combination with chemotherapy or
with chemotherapy and radiation for individuals with carcinoma. Because
chemotherapy and radiation target rapidly growing cells, these
treatments would be more likely to cause a turnover of the colonic
epithelium driving a more significant change in the structure of the
microbiota. Although, we were able to test for an effect across these
specific types of treatment, the number of patients in each treatment
group was relatively small. Finally, those undergoing surgery would have
received antibiotics and this may be a potential confounder. However,
our pre-treatment stool samples were obtained before the surgery and the
post-treatment samples were obtained long after any effects due to
antibiotic administration on the microbiome would be expected to occur
(351 \(\pm\) 102 days).

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}24{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between one and two weeks after initial colonoscopy. The bacterial
community has been shown to normalize back to a pre-colonoscopy
community within this time period {[}27{]}. Our study cohort consisted
of 67 individuals with an initial sample as described and a follow up
sample obtained between 188 - 546 days after treatment of lesion
{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
histopathological review of any biopsies taken. Patients were classified
as having advanced adenoma if they had an adenoma greater than 1 cm,
more than three adenomas of any size, or an adenoma with villous
histology. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}28{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between four sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}28{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}29{]}, screening for
chimeras using UCHIME {[}30{]}, classifying with a naive Bayesian
classifier using the Ribosomal Database Project (RDP){[}31{]}, and
clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm
{[}32{]}. The number of sequences for each sample was rarefied to 10523
to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
used to create the three models used. The adenoma model classified
normal versus adenoma, the advanced adenoma model classified normal
versus advanced adenoma, and the carcinoma model classified normal
versus carcinoma. The total number of individuals in the treatment
models was 67 individuals. There were a total of 22 individuals in the
treatment adenoma model, 19 individuals in the treatment advanced
adenoma model, and 26 individuals in the treatment carcinoma model
{[}Table 1{]}. The total number of individuals in the normal versus
diagnosis models was 423 individuals {[}Table 2{]}. There were a total
of 239 individuals in the adenoma model, 262 individuals in the advanced
adenoma model, and 266 individuals in the carcinoma model {[}Table 2{]}.
Each model was then applied to our 67-person cohort {[}Table 1{]} which
assesed the prediction of pre-treatment adenoma (adenoma n = 22 and
disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
disease free n = 22), pre-treatment advanced adenoma (advanced adenoma n
= 19 and disease free n = 0 ) versus post-treatment advanced adenoma
(advanced adenoma n = 0 and disease free n = 19), pre-treatment
carcinoma (carinoma n = 26 and disease free n = 0) versus post-treatment
carcinoma (carcinoma n = 1 and disease free n = 25).

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}34{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre and post-treatment samples utilized a paired
Wilcoxon ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}35{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtaining the median
MDA from the 100, 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\begin{longtable}[]{@{}lccc@{}}
\toprule
& Adenoma & Advanced Adenoma & Carcinoma\tabularnewline
\midrule
\endhead
n & 22 & 19 & 26\tabularnewline
Age (Mean ± SD) & 61.68 ± 7.2 & 63.11 ± 10.9 & 61.65 ±
12.9\tabularnewline
Sex (\%F) & 36.36 & 36.84 & 42.31\tabularnewline
BMI (Mean ± SD) & 26.86 ± 3.9 & 25.80 ± 4.7 & 28.63 ± 7.2\tabularnewline
Caucasian (\%) & 95.45 & 84.21 & 96.15\tabularnewline
Post-Treatment Sample (Mean ± SD) & 255.41 ± 42 & 250.16 ± 41 & 350.85 ±
102\tabularnewline
\bottomrule
\end{longtable}

@@

\textbf{Figure S1: ROC Curves of the Adenoma, Advanced Adenoma, and
Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
represent the range of values of a 100 different 80/20 splits of the
test set data and the dark green line represents the model using 100\%
of the data set and what was used for subsequent classification. B)
Advanced Adenoma ROC curve: The light yellow shaded areas represent the
range of values of a 100 different 80/20 splits of the test set data and
the dark yellow line represents the model using 100\% of the data set
and what was used for subsequent classification. C) Carcinoma ROC curve:
The light red shaded areas represent the range of values of a 100
different 80/20 splits of the test set data and the dark red line
represents the model using 100\% of the data set and what was used for
subsequent classification.

\textbf{Figure S2: Summary of Important OTUs for the Adenoma, Advanced
Adenoma, and Carcinoma Models.} A) MDA of the most important variables
in the adenoma model. The dark green point represents the mean and the
lighter green points are the value of each of the 100 different runs. B)
Summary of Important Variables in the advanced adenoma model. MDA of the
most important variables in the SRN model. The dark yellow point
represents the mean and the lighter yellow points are the value of each
of the 100 different runs. C) MDA of the most important variables in the
carcinoma model. The dark red point represents the mean and the lighter
red points are the value of each of the 100 different runs.

\textbf{Figure S3: Pre and Post Treatment Relative Abundance of CRC
associated OTUs within the Carcinoma Model.}

@@

\hypertarget{ref-obrien_impact_2013}{}
27. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

\hypertarget{ref-kozich_development_2013}{}
28. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology. 2013;79:5112--20.

\hypertarget{ref-pruesse_silva_2007}{}
29. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
SILVA: A comprehensive online resource for quality checked and aligned
ribosomal RNA sequence data compatible with ARB. Nucleic Acids Research.
2007;35:7188--96.

\hypertarget{ref-edgar_uchime_2011}{}
30. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
sensitivity and speed of chimera detection. Bioinformatics (Oxford,
England). 2011;27:2194--200.

\hypertarget{ref-wang_naive_2007}{}
31. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
for rapid assignment of rRNA sequences into the new bacterial taxonomy.
Applied and Environmental Microbiology. 2007;73:5261--7.

\hypertarget{ref-schloss_assessing_2011}{}
32. Schloss PD, Westcott SL. Assessing and improving methods used in
operational taxonomic unit-based approaches for 16S rRNA gene sequence
analysis. Applied and Environmental Microbiology. 2011;77:3219--26.

\hypertarget{ref-breiman_random_2001}{}
33. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
{[}cited 2013 Feb 7{]};45:5--32. Available from:
\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}

\hypertarget{ref-anderson_permanova_2013}{}
34. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
the face of heterogeneous dispersions: What null hypothesis are you
testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
5{]};83:557--74. Available from:
\url{http://doi.wiley.com/10.1890/12-2010.1}

\hypertarget{ref-benjamini_controlling_1995}{}
35. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.
